Data Specialization for a Staged Lambda Calculus

### 2014-11-04 paper/splitting.tex
rewrote 4.2 with nico

\TODO
The contract for splitting stage \bbone\ terms is more complex, because stage \bbone\ 
terms reduce to a pair of a future context and a partial value, not simply to a value.

%while maintaining the term's original structure.

Stage \bbtwo\ values can contain no stage \bbone\ subcomputations, but the
reverse is not true of partial values (at stage \bbone). These \emph{can}
contain stage \bbtwo\ computations, both inside $\next$ blocks, which must
contain a variable reference to some future context, and inside lambdas, which
may contain a $\next$ in their body.

Thus, to split partial values, we must separate the contents of $\next$ blocks
from the rest of the term. This is implemented by the judgment $v \vsplito
[v_1,v_2]$ (\ref{fig:valSplit}).

No dependence here.

For example, the stage \bbone\ term
\begin{lstlisting}
1`inl(3, next{`2`yhat`1`})`
\end{lstlisting}
splits into
\verb|inl(3,())| and \texttt{((),$\mathtt{\hat{y}}$)}.

sum types

In general,
at product and recursive types, splitting acts
recursively, gathering all subterms' stage $w$ content into $v_w$. At base
types, splitting assigns $v$ entirely to stage \bbone\ (and \texttt{()} to stage
\bbtwo); conversely, at $\fut$ types, splitting assigns $v$ (which must have the
form $\next~\hat y$) entirely to stage \bbtwo\ (and \texttt{()} to stage
\bbone).

This process of exposing only the relevant structure of a term at each stage is
called \emph{masking}.

Splitting a function is more complex, because function bodies are open terms,
and so are split by the judgments described in the next subsection. Essentially,
functions will split into two functions, whose bodies correspond to the stage
\bbone\ and stage \bbtwo\ components of the original function body.
Because that body may contain a precomputation, the generated function signatures must properly emit and resume with that precomputation.

%As described above, splitting a stage \bbtwo\ value always produces a trivial
%precomputation---either \texttt{()}, or nested tuples thereof.

%Thus, it is meaningful to define some notion of splitting for them.  This is
%useful to study, because the definition will later be used as a contract with
%which general stage \bbone\ term splitting must comply.

%Recall that a partial value is externally a stage \bbone\ thing; it has a type
%$@~\bbone$, and along with a future context, it is the result of reducing a
%stage \bbone\ term. But internally, a partial value might contain stage \bbtwo\
%subexpressions.

\crem{value splitting figure is missing injections}

\TODO
\begin{enumerate}
\item Examples
\item Lifting (and masking)
\item Correctness? (Does this require partial value splitting?)
\end{enumerate}

\begin{theorem}[Correctness of splitting at $\bbone$]
If $\vdash e:A~@~\bbone$, then (by splitting)
\begin{itemize}
\item $\vdash e : A \splitonesym [c,l.r]$
\item $\reduce {c} {(v_1,u)}$
\item $\reduce {(\letin{l}{u}{r})} v_S$
\end{itemize}
if and only if (by the staged dynamic semantics)
\begin{itemize}
\item $\diaone [] e {\xi;v}$
\item $\dom \xi \vdash v \vsplito [v_1,v_2]$
\item $\reify \xi{v_2}q$
\item $q \mathbin{\tworedsym} v_D$
\end{itemize}
and if so, then $v_D \sim v_S$.
\end{theorem}

Lifting and masking are two necessary ingredients for ensuring that splitting
observes the stages of \lang. When splitting a partial value, the stage \bbone\
result must \emph{mask} the stage \bbtwo\ computations in order to be
single-staged. When splitting a stage \bbtwo\ term, the precomputation must
\emph{lift} out all embedded stage \bbone\ computations, so they can be
performed first, even though their results are only used in stage \bbtwo.
%and it's accessible to whatever surrounding piece of code refers to the value being split.
%is not accessible externally.

Splitting a stage \bbone\ term, in contrast, requires both masking and lifting.
Consider the term 
\begin{lstlisting}
1`(next{`2`hold{`1` 1+2 `2`} < 5`1`},3*4)`
\end{lstlisting}
Masking yields the stage \bbone\ computation \verb|((),3*4)|, while lifting
yields the (stage \bbone) precomputation \verb|(1+2)|. While both computations
must occur at stage \bbone, the former is needed by stage \bbone, while the
latter is needed by stage \bbtwo.

The precomputation for this term is a \emph{combined result} which is
the pair of these components: \verb|(((),3*4),1+2)|. The resumer is
likewise the pair of the stage \bbtwo\ results of lifting and masking:
\verb|fn L => (L<5,())|.
%\footnote{In this example, the combined result is a 2-tuple, but more generally
%it only must be a term which reduces to a 2-tuple.}

Splitting is implemented by the judgment $\splitonesym$,
the rules for which are given in \cref{fig:termSplitOne}.  
It operates inductively on the structure of the term.  
The $\splitonesym$ judgment sends a stage \bbone\ term $e$ to a combined term
$c$ and resumer $l.r$.

@@

>>>>

Splitting a stage \bbone\ term $\colone e A$ produces a term and a function,
which respectively contain all the stage \bbone\ and stage \bbtwo\
subcomputations of $e$. The situation is more complex than when splitting stage
\bbtwo\ terms; to see why, consider the correctness criterion for splitting
stage \bbone\ terms.

Stage \bbone\ evaluation of $e$, via $e\mathbin{\redonesym}[\xi;v]$,
yields a table of (stage \bbtwo) residuals $\xi$, and a (stage \bbone) partial
value $v$. It is tempting to compare the two results of evaluating $e$ to the
two results of splitting it. To see where this goes wrong, consider the stage
\bbone\ term
\begin{lstlisting}
1`(1+2, next{`2`hold{`1`3+4`2`} + 5`1`})`
\end{lstlisting}
which evaluates to
\begin{lstlisting}
2`[yhat|->7+5]` 1`(3,next{`2`yhat`1`})`
\end{lstlisting}
This partial value contains a $\next$, and so is itself multistaged:
intuitively, its stage \bbone\ part is \texttt{3}, while its stage \bbtwo\ part
is $\mathtt{\hat{y}}$, which is bound to \texttt{7+5}.

We formalize this notion via two \emph{masking} functions (\ref{fig:valSplit}),
$\masko{\cdot}$ and $\maskt{\cdot}$, which expose only the relevant structure of
a partial value at each stage, essentially by replacing all stage \bbone\
(resp., \bbtwo) components of a partial value by \texttt{()}. The first result
of splitting must then compute the \bbone-mask of the partial value, and the
second result the \bbtwo-mask.

There is one final subtlety to splitting stage \bbone\ terms. In the example
above, there are \emph{two} flavors of stage \bbone\ subcomputations:
\texttt{1+2}, which is needed to compute the \bbone-mask, and \texttt{3+4},
a precomputation which is used by the computation of the \bbtwo-mask.

Formally, splitting a stage \bbone\ term, as implemented by the judgment $e
\splitonesym [c,l.r]$ (\ref{fig:termSplitOne}), yields a \emph{combined result}
$c$ which computes a pair of the stage \bbone\ result and the precomputation,
and a \emph{resumer} function $l.r$. The correctness of splitting requires that,
for $e\mathbin{\redonesym}[\xi;v]$, we have that $\pio~c$ agrees with
$\masko{v}$, and $\letin l{\pit~c}r$ agrees with $\maskt{\xi;v}$.

The example splits into a combined result $c=$ \texttt{((1+2,()), 3+4)}, and a
resumer $l.r=$ \texttt{L => ((),L+5)}. Then $\pio~c$ reduces to \texttt{(3,())},
which agrees with the stage \bbone\ mask, and $\letin l{\pit~c}r$ reduces to
\texttt{((),12)}, which agrees with the stage \bbtwo\ mask.

\paragraph{Stage \bbone\ masking}

The purpose of masking is to expose only the relevant structure of a partial
value $v$ at each stage. For the most part, stage \bbone\ masking, $\masko{v}$,
simply replaces all $\next$s in $v$ with \texttt{()}. The only exception is for
functions $\lam xAe$; such a value represents a computation, so masking splits
its body as a computation, $e\splitonesym [c,l.r]$, and replaces it with its
stage \bbone\ component, the combined result.

\paragraph{Stage \bbtwo\ masking}

Stage \bbtwo\ masking, unlike stage \bbone\ masking, requires access to the
future context $\xi$, because the stage \bbtwo\ parts of $v$ contain variables
in $\xi$. To compute $\maskt{\xi;v}$, we rewrite $v$ to expose only the relevant
structure, and then reify the resulting term in the context $\xi$. This
rewriting process replaces base type constants in $v$ with \texttt{()}, and
$\next~\hat{y}$ with $\hat{y}$. Injection tags are eliminated, because the sum
type is at stage \bbone, so the tag is not relevant at stage \bbtwo. For
functions $\lam xAe$, masking again splits its body $e\splitonesym [c,l.r]$,
replaces it with the resumer, and uncurries, to yield $\lambda (x,l).r$.

\paragraph{Splitting}

Splitting stage \bbone\ terms, as implemented by the judgment $e\splitonesym
[c,l.r]$, is also guided by the structure of the term $e$.

\TODO polish the following explanation

@@

\crem{split value splitting into two judgments, and call it masking}

\crem{masking figure is missing injections}

\crem{rename future context to residual table}
